---
title: "Exploratory analysis Flanders double-sampling study"
format: 
  html:
    embed-resources: true
toc: true
---

This short report contains my exploratory work with the Flanders double-sampling study and contains some questions on how the data should be used.

```{r setup, echo = F, warning = F, messages = F}

library(tidyverse) |> suppressPackageStartupMessages()
library(here) |> suppressPackageStartupMessages()
library(sf) |> suppressPackageStartupMessages()
library(inbospatial) |> suppressPackageStartupMessages()
library(units) |> suppressPackageStartupMessages()
```

## Data import and inspection

The dataset contains three csv files.

### (1) Sampling points

```{r data-import1}

folder_flds <- here("data", "Flanders_Double_Sampling_Data")

# sampling points
sampling_points <- read_delim(
  here(folder_flds,
       "EJPsoil_LUCAS22_Double_sampling_BE_Flanders_Sampling_Points_list.csv"),
  delim = ";",
  col_types = cols(
    # import as characters to fix issues with different notation:
    # both . and , are used as decimal character ???
    WGS84Latitude = col_character(),
    WGS84Longitude = col_character(),
    Elevation = col_character()
  )
)
glimpse(sampling_points)
```

Little data cleaning and formatting as `sf` object:

```{r data-cleaning1}
sampling_points_clean <- sampling_points %>%
  mutate(
    across(c(WGS84Latitude, WGS84Longitude, Elevation), 
           ~ as.numeric(gsub(",", ".", .x)))
  ) %>%
  st_as_sf(coords = c("WGS84Longitude", "WGS84Latitude"),
           crs = "WGS84") %>%
  select(LUCAS_PlotID, Town, Elevation, contains("LU"), RSG, Bio_Point)

glimpse(sampling_points_clean)
```

Five data points lie in Wallonia and 1 in Brussels. Question: Do we need to retain them for the comparison of sampling schemes (LUCAS vs. Cmon)?

```{r data-cleaning2}

# import provinces of Flanders
ogc_vrbg <- "https://geo.api.vlaanderen.be/VRBG2025/ogc/features/v1/"
provinces <- get_feature_ogc(
  url = ogc_vrbg,
  collection = "Refprv",
  crs = "EPSG:31370"
) %>% st_transform(crs = "WGS84")

# print points outside Flanders
outside_idx <- st_disjoint(
  sampling_points_clean, 
  st_union(provinces), 
  sparse = FALSE
)
sampling_points_clean[outside_idx, ] %>%
  st_drop_geometry() %>%
  select(Town) %>%
  print()

# filter on points inside Flanders (N = 160)
sampling_points_fl <- st_filter(
  sampling_points_clean,
  provinces
)
ggplot() +
  geom_sf(data = provinces) +
  geom_sf(data = sampling_points_fl)
```

### (2) Soildata

```{r data-import2, message = F}
soildata <- read_csv2(
  here(folder_flds,
       "EJPsoil_LUCAS22_Double_sampling_BE_Flanders_Soildata.csv")
)
glimpse(soildata)
```

Little data cleaning step and filtering observations. Questions/remarks: Do we need to retain the data from the strooisel-laag? There is 1 plot with 6 observations, are these partial duplicates? There are two points that are listed in the previous csv file but have no corresponding records in the soildata csv.

```{r data-cleaning3}
# Depth data were formatted as a date (probably a setting from Excel/GSheet?)
soildata_clean <- soildata %>%
  mutate(
    Depth = case_when(Depth == "okt/30" ~ "10-30", 
                      TRUE ~ Depth)
  )

# 136 plots have 3 observations (2*Cmon + 1*LUCAS), 29 plots have 5 observations (mineral + organic matrix) and 1 plot has 6 observations (partial duplicates?)
soildata_clean %>%
  summarise(.by = LUCAS_PlotID, N = n()) %>%
  summarise(.by = N, plots = n())

# selecting soildata: remove strooisel-laag, keep only points in Flanders and remove partial duplicates
soildata_clean_fl <- soildata_clean %>%
  filter(
    !grepl("strooisel", tolower(FieldSampleID)),
    LUCAS_PlotID %in% sampling_points_fl$LUCAS_PlotID
  ) %>%
  distinct(LUCAS_PlotID, Depth, .keep_all = T)

# there are 2 points that appear in the sampling_points object but that don't appear in the soildata:
sampling_points_fl %>%
  filter(!(LUCAS_PlotID %in% soildata_clean_fl$LUCAS_PlotID)) %>%
  st_drop_geometry() %>%
  select(LUCAS_PlotID, Town) %>%
  print()
```

Question: Why are some TOC values sometimes larger than TC? (is TC not the sum of TOC and TIC?; is there a rounding issue?) Missing values in TIC when TC and TOC are both observed, can you just use TC - TOC as imputed value? 

```{r question-TC}
soildata_clean_fl %>%
  mutate(
    TC  = janitor::round_half_up(TC, 1),
    TOC = janitor::round_half_up(TOC, 1),
    TIC = janitor::round_half_up(TIC, 1)
  ) %>%
  filter(TOC > TC) %>%
  select(LUCAS_PlotID, TC, TOC, TIC)
```

Question: To compare LUCAS vs. Cmon, should we combine the two Cmon measurements? Can you take the weighted average of the 0-10 cm sample and the 10-30 cm sample (with respective weights 1/3 and 2/3) to get an average for the 0-30 cm depth?  

```{r question-Cmon-avg}
# weighted average for Cmon sampling scheme:
soildata_clean_fl_avg <- soildata_clean_fl %>%
  mutate(
    weights = case_when(Depth == "0-10"  ~ 1/3,
                        Depth == "10-30" ~ 2/3,
                        Depth == "0-30"  ~ 1),
    .after = Depth
  ) %>%
  summarise(
    .by = c(LUCAS_PlotID, SampScheme, Matrix),
    Depth = "0-30",
    Thickness = sum(Thickness),
    across(c(Sand, Silt, Clay, TC, TIC, TOC, TN, 
             contains("pH"), contains("EC")), 
           ~ weighted.mean(.x, w = weights, na.rm = T))
  ) %>%
  mutate(
    across(c(Sand, Silt, Clay, TC, TIC, TOC, TN, 
             contains("pH"), contains("EC")),
           ~ replace_when(.x, is.nan(.x) ~ NA_real_))
  )
```

### (3) Bulk density

```{r data-import3, message = F}
bulkdens <- read_csv2(
  here(folder_flds,
       "EJPsoil_LUCAS22_Double_sampling_BE_Flanders_Bulk_density.csv")
)
glimpse(bulkdens)
```

Little data cleaning step and weighted-average:

```{r data-cleaning4}
bulkdens_clean <- bulkdens %>%
  mutate(
    across(c(contains("BD_"), contains("SWC_")), ~ as.numeric(.x))
  ) %>%
  mutate(
    weights = case_when(Depth == "0-10"  ~ 1/3,
                        Depth == "10-30" ~ 2/3,
                        Depth == "0-30"  ~ 1),
    .after = Depth
  ) %>%
  summarise(
    .by = LUCAS_PlotID,
    Depth = "0-30",
    SampScheme = "Cmon",
    across(c(contains("BD_"), contains("SWC_")), 
           ~ weighted.mean(.x, w = weights, na.rm = T))
  )
```

### Combine datasets and some further cleaning

```{r combine-datasets}
soildata_flds <- soildata_clean_fl_avg %>%
  left_join(sampling_points_fl, by = "LUCAS_PlotID") %>%
  left_join(bulkdens_clean, by = c("LUCAS_PlotID","Depth","SampScheme")) %>%
  select(
    LUCAS_PlotID, Town, Elevation, geometry, LUCAS_LU, FieldObs_LU, RSG, 
    #FieldSampleID, LabSampleID, 
    SampScheme, Depth, Matrix,       
    Thickness, Sand, Silt, Clay, TC, TIC, TOC, TN, 
    contains("pH_"), contains("EC_"),
    #LayerID, 
    BD_mean, BD_sd, BD_CV, SWC_volp_mean, SWC_volp_sd
  ) %>%
  mutate(
    Thickness = set_units(Thickness, "cm"),
    Elevation = set_units(Elevation, "m"),
    Sand = set_units(Sand, "%"),
    Silt = set_units(Silt, "%"),
    Clay = set_units(Clay, "%"),
    TC = set_units(TC, "g/kg"),
    TIC = set_units(TIC, "g/kg"),
    TOC = set_units(TOC, "g/kg"),
    TN = set_units(TN, "g/kg"),
    EC_m_v = set_units(EC_m_v, "µS/cm"),
    EC_v_v = set_units(EC_v_v, "µS/cm"),
    BD_mean = set_units(BD_mean, "g/cm³"),
    BD_sd = set_units(BD_sd, "g/cm³"),
    BD_CV = set_units(BD_CV, "g/cm³"),
    SWC_volp_mean = set_units(SWC_volp_mean, "%"),
    SWC_volp_sd = set_units(SWC_volp_sd, "%")
  )

summary(soildata_flds)
```

## Data exploration

### Landuse & soil types

Subgroup analyses within land-use and soil types will need to be restricted to groups with enough observations. 

```{r}
soildata_flds %>%
  summarise(.by = FieldObs_LU, N = n_distinct(LUCAS_PlotID)) %>%
  arrange(desc(N)) %>%
  print(n = 5)

soildata_flds %>%
  summarise(.by = RSG, N = n_distinct(LUCAS_PlotID)) %>%
  arrange(desc(N)) %>%
  print(n = 15)
```

### Comparing LUCAS & Cmon sampling schemes

The graphs below show a scatter plot and loess smoother for the parameter measured with LUCAS vs. Cmon (weighted average). The bisector is added for reference. Detection of bivariate outliers is performed with the robust method Minimum Covariance Determinant (`performance` package). Not all the detected outliers will be influential for model fitting but may signal interesting differences between LUCAS and Cmon. Alternative: outlier detection with Bland-Altman analysis.

```{r comparing-lucas-cmon, warning = F, message = F}

# plot + outlier detection with Minimum Covariance Determinant
compare_schemes <- function(vrb) {
  data_wide <- soildata_flds %>%
    select(LUCAS_PlotID, SampScheme, all_of(vrb)) %>%
    pivot_wider(names_from = SampScheme, values_from = all_of(vrb),
                names_prefix = paste0(vrb, "_")) %>%
    filter(
      !is.na(.data[[paste0(vrb,"_Cmon")]]),
      !is.na(.data[[paste0(vrb,"_LUCAS")]])
    ) 
  outl <- performance::check_outliers(
    data_wide %>% 
      select(-LUCAS_PlotID), 
    method = "mcd", percentage_central = .75,
    verbose = TRUE
  )
  data_wide$outlier <- FALSE
  data_wide$outlier[which(outl)] <- TRUE
  
  data_wide %>%
    mutate(
      lab_outl = case_when(outlier ~ LUCAS_PlotID, TRUE ~ NA_real_)
    ) %>%
    ggplot(aes(.data[[paste0(vrb,"_Cmon")]], 
               .data[[paste0(vrb,"_LUCAS")]], 
               color = outlier)) +
    geom_point() +
    geom_smooth(
      inherit.aes = F,
      aes(.data[[paste0(vrb,"_Cmon")]], 
               .data[[paste0(vrb,"_LUCAS")]]),
      method = "loess", color = "grey"
    ) +
    ggrepel::geom_text_repel(aes(label = lab_outl), show.legend = F,
                             size = 2) +
    geom_abline(slope = 1, intercept = 0, lty = 2)
}

# apply function
sapply(
  FUN = compare_schemes,
  X = c("Sand","Silt","Clay","TC","TOC","TIC","TN",
        "pH_KCl_v_v","pH_KCl_m_v",
        "pH_H20_m_v","pH_H20_v_v",
        "pH_CaCl2_m_v","pH_CaCl2_v_v",
        "EC_m_v","EC_v_v")
)
```

Question: For the comparison between Cmon and LUCAS; should a TF be developed to translate (predict) values between them or are descriptive statistics enough? Eg. Bland-Altman, Wilcox-test.

```{r wilcox-Cmon-LUCAS}
# to be decided: statistical comparison with a Wilcox test (location-shift)
# => need to apply a multiple-comparison correction (15 variables tested)
compare_schemes_wilcox <- function(vrb) {
  data_wide <- soildata_flds %>%
    select(LUCAS_PlotID, SampScheme, all_of(vrb)) %>%
    mutate(across(all_of(vrb), ~ as.numeric(.x))) %>%
    pivot_wider(names_from = SampScheme, values_from = all_of(vrb),
                names_prefix = paste0(vrb, "_")) %>%
    filter(
      !is.na(.data[[paste0(vrb,"_Cmon")]]),
      !is.na(.data[[paste0(vrb,"_LUCAS")]])
    )
  wilcox.test(
    data_wide[[paste0(vrb,"_Cmon")]],
    data_wide[[paste0(vrb,"_LUCAS")]],
    paired = T, conf.int = T, correct = T,
  ) %>% broom::tidy()
}
  
# apply function
sapply(
  FUN = compare_schemes_wilcox,
  X = c("Sand","Silt","Clay","TC","TOC","TIC","TN",
        "pH_KCl_v_v","pH_KCl_m_v",
        "pH_H20_m_v","pH_H20_v_v",
        "pH_CaCl2_m_v","pH_CaCl2_v_v",
        "EC_m_v","EC_v_v")
) %>% t() %>%
  as.data.frame() %>%
  select(estimate, p.value, conf.low, conf.high) %>%
  # FDR correction
  mutate(q.value = p.adjust(p.value, method = "BH")) %>%
  print(digits = 2)
```

### Transfer functions pH

For this analysis we don't need the averaged Cmon data because we need to compare methods that are measured at each depth, but depth could play the role of a covariate if necessary. Strong correlations (>= .98) are observed between all pairs of pH measurement methods (same when Spearman instead of Pearson).

```{r scatmat, warning = F}
soildata_clean_fl %>%
  select(contains("pH")) %>%
  GGally::ggscatmat(corMethod = c("pearson"))
```

